# Variability in activity patterns {#var}

To illustrate how site-to-site variability can affect the estimate of activity patterns, we simulated data under equation 1 explained in the main text and reported here:

$$
y_{it} \sim Bernoulli(p_{it})
$$

$$
logit(p_{it}) = \beta_0 + \beta_1*cos(\frac{2\pi t}{24} + (\theta_0 + \gamma_i)) + \beta_2*cos(\frac{2\pi t}{12} + (\theta_1 + \gamma_i)) + \tau_i
$$

$$
\tau_i \sim N(0,\sigma_\tau)
$$

$$
\gamma_i \sim N(0,\sigma_\gamma)
$$

We load the necessary libraries and set the intercept $\beta_0$, and the two amplitudes $\beta_1$ and $\beta_2$ equal to -3, 1, and 0.7, respectively. We assign the values 3 and 2 to the common phaseshifts terms, $\theta_0$ and $\theta_1$, respectively. 

```{r sim1, message = FALSE, warnings = FALSE}
# Load libraries and set equations' parameters
set.seed(129)
library(tidyverse)
library(grid)
library(gridExtra)
wavelength = 24
b0 = -3
b1 = 1 
b2 = 0.7
theta0 = 3
theta1 = 2 
tau_i = gamma_i = seq(-1.5, 1.5, by = 0.5)
time <- seq(0, 23, length.out = 100)
```

## Variability in frequency of site-use

To create examples of *variability in frequency of site-use*, we keep all the parameters constant (with $\gamma_i = 0$) except for $\tau_i$, which varies from -1.5 to 1.5 in increments of 0.5. For illustrative purposes, we assign these predefined values to $\tau_i$, instead of drawing random values from the distribution $N(0,\sigma_\tau)$. We create the curves in logit scale but trasform them in the probability scale, for easiness in interpretation. We also include code to calculate conditional and marginal mean activity curves (red and black curves in the following plots); we cover this aspect in detail in section \@ref(condvsmarg) and in Box 1 in the main text. Here, we briefly say that the conditional mean curve describes the activity pattern at a typical site, whereas the marginal mean curve describes the population-average mean, the activity patterns averaged across all the sites.  

```{r sim2}
# create the curves
p_df <- data.frame(p = c(
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[1]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[2]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[3]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[5]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[6]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[7])
), time = rep(time, 7), 
mean = c(rep(rep("off", length(time)),3), rep("on", length(time)), rep(rep("off", length(time)),3)),
curvesID = as.factor(rep(1:7, each = length(time))),
curvesLeg = as.factor(rep(c("+/- 1.5 SD", "+/- 1 SD", "+/- 0.5 SD", "Conditional", "+/- 0.5 SD", "+/- 1 SD", "+/- 1.5 SD"), each = length(time)))
)

# calculate the marginal mean activity curve
p_df2 <- p_df %>% select(p, time) %>% group_by(time) %>% 
  summarize(p = mean(p)) %>% 
  mutate(mean = "on", curvesID = as.factor(8),
         curvesLeg = as.factor("Marginal"))

p_df <- rbind(p_df, as.data.frame(p_df2))

# plot
(pl_vert <- ggplot(p_df, aes(x = time, y = p, group = curvesID)) +
  geom_line(aes(color = curvesLeg, linewidth = mean, linetype = mean)) +
  scale_color_manual(values = c("grey40", "grey60", "grey80", "red", "black")) +
  scale_linewidth_manual(values = c(0.5, 1.5)) +
  scale_linetype_manual(values = c(2, 1, 2)) +
  coord_cartesian(ylim = c(-0.05, 0.40)) +
  labs(x = "Time of Day", y = "Probability of Activity", title = "A) Variability in \n frequency of site-use") +
  theme_minimal()+
  theme(legend.position = "none",
        axis.line = element_line(colour = 'black', linetype = 'solid'),
        axis.ticks.x = element_line(colour = 'black', linetype = 'solid'), #element_blank(), #
        axis.text.y = element_blank(), 
        axis.title=element_text(size = 8,face = "bold"),
        plot.title=element_text(size = 9,face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) + 
  geom_segment(aes(x=12, y=min(p), xend=12, yend=min(p)-0.05), arrow=arrow(length= unit(0.5, "cm")), 
               linewidth = 1, color = "orange", linejoin = "mitre") +
  geom_segment(aes(x=12, y=0.25+0.003, xend=12, yend=0.25+0.053), arrow=arrow(length= unit(0.5, "cm")), 
               linewidth = 1, color = "orange", linejoin = "mitre")+
  scale_x_continuous(breaks=seq(0,24,length.out=7), labels=seq(0,24,4))) 

```

In this first plot, the dashed grey curves represent the curves generated using the different values assigned to $\tau_i$ (darker to lighter colors from lower to higher in absolute values). All other things equal, these different values of $\tau_i$ lead to a vertical shift in the activity curves, which corresponds to a change in the frequency of site-use. 

## Variability in timing of activity peak

To create examples of *variability in timing of activity peak*, we set the random phaseshift $\gamma_i$ as equal to 0, $\pm$ 0.5, $\pm$ 1.0, and $\pm$ 1.5, while keeping $\tau_i = 0$ and the remaining parameters with the same values used before. We create and plot the curves. Then, we joint the two plots to created the first portion of Figure 1, reported in the main text. 

```{r sim3}
# create the curves
p_df <- data.frame(p = c(
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[1]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[1]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[2]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[2]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[3]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[3]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[4]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[5]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[5]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[6]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[6]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[7]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[7]) + tau_i[4])
), time = rep(time, 7), 
mean = c(rep(rep("off", length(time)),3), rep("on", length(time)), rep(rep("off", length(time)),3)),
curvesID = as.factor(rep(1:7, each = length(time))),
curvesLeg = as.factor(rep(c("+/- 0.5 SD", "+/- 1 SD", "+/- 1.5 SD", "Conditional", "+/- 0.5 SD", "+/- 1 SD", "+/- 1.5 SD"), each = length(time))) 
)

# calculate the marginal mean activity curve
p_df2 <- p_df %>% select(p, time) %>% group_by(time) %>% 
  summarize(p = mean(p)) %>% 
  mutate(mean = "on", curvesID = as.factor(8),
         curvesLeg = as.factor("Marginal"))

p_df <- rbind(p_df, as.data.frame(p_df2))

# plot
pl_hor <- ggplot(p_df, aes(x = time, y = p, group = curvesID)) +
  geom_line(aes(color = curvesLeg, linewidth = mean, linetype = mean)) +
  scale_color_manual(values = c("grey40", "grey60", "grey80", "red", "black")) +
  scale_linewidth_manual(values = c(0.5, 1.5)) +
  scale_linetype_manual(values = c(2, 1)) +
  labs(x = "Time of Day", y = "Probability of Activity", title = "B) Variability in \n timing of activity peaks") +
  theme_minimal()+
  theme(legend.position = "none", 
        axis.line = element_line(colour = 'black', linetype = 'solid'),
        axis.ticks.x = element_line(colour = 'black', linetype = 'solid'), 
        axis.text.y = element_blank(), 
        axis.title=element_text(size=8,face="bold"),
        plot.title=element_text(size=9,face="bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  geom_segment(aes(x=12 - 0.75, y=max(p) + 0.02, xend=12 - 7, yend=max(p) + 0.02), arrow=arrow(length= unit(0.5, "cm")), 
               linewidth = 1, color = "orange", linejoin = "mitre") +
  geom_segment(aes(x=12 + 0.75, y=max(p) + 0.02, xend=12 + 7, yend=max(p) + 0.02), arrow=arrow(length= unit(0.5, "cm")), 
               linewidth = 1, color = "orange", linejoin = "mitre") +
  scale_x_continuous(breaks=seq(0,24,length.out=7), labels=seq(0,24,4))+
  guides(linetype=element_blank())


# Join plots
pl <- grid.arrange(pl_vert, pl_hor, ncol = 2)

```

The dashed lines in panel B represent curves simulated using different values of $\gamma_i$ (darker to lighter colors from lower to higher in absolute values). Varying the random phaseshift $\gamma_i$ results in activity curves that are shifted horizontally. Because we are simulating data using two cosine terms, the cumulative effects of the two random phaseshifts (one per cosine term) also affect the intensity of site-use. Again, we will discuss in more details when illustrating the differences between conditional and marginal mean activity curves in section \@ref(condvsmarg). 

For completeness, we also present the example of a unimodal pattern. We can simulate this activity pattern by re-running the code above, this time setting $\beta_2 = 0$ (see main text). Setting $\beta_2 = 0$ eliminates the second cosine term and returns a curve with only one peak in activity within the 24-hour cycle.

```{r sim4}
# setting beta2 equal to zero 
b2 = 0

# change in frequency of site-use
# create the curves
p_df <- data.frame(p = c(
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[1]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[2]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[3]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[5]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[6]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+theta0 + gamma_i[4]) + tau_i[7])
), time = rep(time, 7), 
mean = c(rep(rep("off", length(time)),3), rep("on", length(time)), rep(rep("off", length(time)),3)),
curvesID = as.factor(rep(1:7, each = length(time))),
curvesLeg = as.factor(rep(c("+/- 1.5 SD", "+/- 1 SD", "+/- 0.5 SD", "Conditional", "+/- 0.5 SD", "+/- 1 SD", "+/- 1.5 SD"), each = length(time)))
)

# calculate the marginal mean activity curve
p_df2 <- p_df %>% select(p, time) %>% group_by(time) %>% 
  summarize(p = mean(p)) %>% 
  mutate(mean = "on", curvesID = as.factor(8),
         curvesLeg = as.factor("Marginal"))

p_df <- rbind(p_df, as.data.frame(p_df2))

# plot
pl_vert_uni <- ggplot(p_df, aes(x = time, y = p, group = curvesID)) +
  geom_line(aes(color = curvesLeg, linewidth = mean, linetype = mean)) +
  scale_color_manual(values = c("grey40", "grey60", "grey80", "red", "black")) +
  scale_linewidth_manual(values = c(0.5, 1.5)) +
  scale_linetype_manual(values = c(2, 1, 2)) +
  coord_cartesian(ylim = c(-0.05, max(p_df$p)+0.053)) +
  labs(x = "Time of Day", y = "Activity Patterns", title = "A) Variability in frequency of site-use") +
  theme_minimal()+
  theme(legend.position = "none",
        axis.line = element_line(colour = 'black', linetype = 'solid'),
        axis.ticks.x = element_line(colour = 'black', linetype = 'solid'), 
        axis.text.y = element_blank(), #element_text(size=8),
        axis.title=element_text(size=10,face="bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) + 
  geom_segment(aes(x=12, y=min(p), xend=12, yend=min(p)-0.05), arrow=arrow(length= unit(0.5, "cm")), 
               linewidth = 1, color = "orange", linejoin = "mitre") +
  geom_segment(aes(x=12, y=max(p)+0.003, xend=12, yend=max(p)+0.053), arrow=arrow(length= unit(0.5, "cm")), 
               linewidth = 1, color = "orange", linejoin = "mitre")+
  scale_x_continuous(breaks=seq(0,24,length.out=7), labels=seq(0,24,4)) 

# change in timing of activity peak
# creating the curves
p_df <- data.frame(p = c(
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[1]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[1]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[2]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[2]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[3]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[3]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[4]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[4]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[5]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[5]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[6]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[6]) + tau_i[4]),
  boot::inv.logit(b0 + b1*cos(2*pi*time/(24)+theta0 + gamma_i[7]) + b2*cos(2*pi*time/(12)+ theta1 + gamma_i[7]) + tau_i[4])
), time = rep(time, 7), 
mean = c(rep(rep("off", length(time)),3), rep("on", length(time)), rep(rep("off", length(time)),3)),
curvesID = as.factor(rep(1:7, each = length(time))),
curvesLeg = as.factor(rep(c("+/- 0.5 SD", "+/- 1 SD", "+/- 1.5 SD", "Conditional", "+/- 0.5 SD", "+/- 1 SD", "+/- 1.5 SD"), each = length(time))) 
)

# calculate the marginal mean activity curve
p_df2 <- p_df %>% select(p, time) %>% group_by(time) %>% 
  summarize(p = mean(p)) %>% 
  mutate(mean = "on", curvesID = as.factor(8),
         curvesLeg = as.factor("Marginal"))

p_df <- rbind(p_df, as.data.frame(p_df2))

# plot
pl_hor_uni <- ggplot(p_df, aes(x = time, y = p, group = curvesID)) +
  geom_line(aes(color = curvesLeg, linewidth = mean, linetype = mean)) +
  scale_color_manual(values = c("grey40", "grey60", "grey80", "red", "black")) +
  scale_linewidth_manual(values = c(0.5, 1.5)) +
  scale_linetype_manual(values = c(2, 1)) +
  labs(x = "Time of Day", y = "Activity Patterns", title = "B) Variability in timing of activity peaks") +
  theme_minimal()+
  theme(legend.position = "none", 
        axis.line = element_line(colour = 'black', linetype = 'solid'),
        axis.ticks.x = element_line(colour = 'black', linetype = 'solid'),
        axis.text.y = element_blank(), #element_text(size=8),
        axis.title=element_text(size=10,face="bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) +
  geom_segment(aes(x=13.5 - 0.75, y=max(p) + 0.02, xend=13.5 - 7, yend=max(p) + 0.02), arrow=arrow(length= unit(0.5, "cm")), 
               linewidth = 1, color = "orange", linejoin = "mitre") +
  geom_segment(aes(x=13.5 + 0.75, y=max(p) + 0.02, xend=13.5 + 7, yend=max(p) + 0.02), arrow=arrow(length= unit(0.5, "cm")), 
               linewidth = 1, color = "orange", linejoin = "mitre") +
  scale_x_continuous(breaks=seq(0,24,length.out=7), labels=seq(0,24,4))+
  guides(linetype=element_blank())


# Join plots
pl_uni <- grid.arrange(pl_vert_uni, pl_hor_uni, ncol = 2)
```
This second example clearly illustrate how varying $\gamma_i$ (while keeping all the other parameters constant) leads to the horizontal shift of the curves, with peaks in activity equal across time but occurring at different times of the diel cycle.

In the next section, we simulate data drawing values of $\tau_i$ and $\gamma_i$ from the normal distributions specified at the beginning of this section. We use this simulated data to illustrate how trigonometric and cyclic cubic regression spline HMs can be used to estimate activity patterns. In section \@ref(condvsmarg), we illustrate the difference between conditional and marginal mean activity patterns and explained how this difference can be relevant when answering ecological questions.
