# Unimodal versus binomial activity {#unibim}

In this first example, we illustrate how trigonometric and cyclic cubic spline GLMMs can 
be used to explore the what diel pattern best describe that of the target species. 
Traditionally, diel activity patterns are usually classified as:
* **unimodal**: the species is mostly active in a certain period of the 24-hour cycle,
and not (or almost not) active in the remaining time. The activity pattern curve
is characterized by only one peak, and it is typically of diurnal and nocturnal species.
* **bimodal**: the species is mostly active during two specific time of 24-hour cycle,
often corresponding to sunrise and sunset (i.e., crepuscular species), and inactive (or mostly inactive) during the 
rest of the time. The activity pattern curve is characterized by having two peaks.
* **cathemeral**: the species is equally active throughout the 24-hour cycle.

We use independent camera-trap records of coyotes (*Canis latrans*) collected at 100 locations between 2016 and 2018 in Northern Minnesota, USA.


## The simplest Trigonometric GLMMs
We use the data as prepared in the previous section, \@ref(dataprep)



You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
