# Evaluating Hypotheses Regarding the Shape of Activity Curves {#unibim}

Traditionally, diel activity patterns have been classified as either:

* **unimodal**: the species is mostly active during a restricted portion of the 24-hour cycle, and not (or almost not) active in the remaining time. The activity  curve is characterized by only one peak, and it is typical of species that are either diurnal or nocturnal.
* **bimodal**: the species is mostly active during two specific times of 24-hour cycle, often corresponding to sunrise and sunset (i.e. crepuscular species) and inactive (or mostly inactive) during the rest of the time. The activity curve is characterized by having two peaks.
* **cathemeral**: the species is equally active throughout the 24-hour cycle and the activity curve approximates a flat line.

Here, we build and compare trigonometric HMs representing these three different diel activity patterns. To reiterate, we use independent camera-trap records of coyotes (*Canis latrans*) collected between 2016 and 2018 at 100 locations in Northern Minnesota, USA, [@Iannarilliea2021], with the data organized as described in the previous chapter \@ref(dataprep). 

<!---JF note to Fabiola below--->
<span style="color: red;">Can we delete this code and just read in the data from the previous chapter?</span>.


``` {r univsbim1, message = TRUE, warnings = TRUE}
# Load libraries
rm(list = ls())
set.seed(129)
library(dplyr)
library(lubridate)
library(GLMMadaptive)
library(lmtest)
library(ggpubr)

# Load data: Coyote 
dat <- read.csv("data_input/species_records.csv") %>% 
  filter(Species == "Coyote") %>% droplevels %>%  
  mutate(DateTimeOriginal = ymd_hms(DateTimeOriginal))
cov <- read.csv("data_input/CameraTrapProject_CT_data_for_analysis_MASTER.csv", as.is = TRUE) %>% 
  select(c(Session, Site, Date_setup, Date_retr, Problem1_from, Problem1_to)) %>% 
  mutate(Date_setup = mdy(Date_setup),
         Date_retr = mdy(Date_retr),
         Problem1_from = mdy(Problem1_from),
         Problem1_to = mdy(Problem1_to)) 

# Merge time of deployment and retrieval + problems
site <- cov
site$end <- ymd("2000-01-01")
for(i in 1:nrow(site)){
  site$end[i] <-  min(site$Date_retr[i], site$Problem1_from[i], na.rm = TRUE)
}


# Create dataframe to store captures 
# (model does not converge if using 30 minutes)
occasions <- vector("list", length = nrow(site))
for(i in 1:nrow(site)){
  occasions[[i]] <- data.frame(Session = site$Session[i],
                               Site = site$Site[i],
                               start = seq(from = ymd_hms(paste(site$Date_setup[i], "00:00:00", sep = " ")), 
                                           to = ymd_hms(paste(site$end[i], "23:59:59", sep = " ")), by = '60 min')) %>% 
    mutate(end = c(start[2:length(start)], start[length(start)]+minutes(60))) 
}
occasions <- do.call(rbind.data.frame, occasions)
occasions$capt <- 0

# Store captures
for(i in 1:nrow(dat)){
  occasions[occasions$Session == as.character(dat$Session[i]) & occasions$Site == as.character(dat$Station[i]) &
              occasions$start <= dat$DateTimeOriginal[i] & occasions$end > dat$DateTimeOriginal[i], "capt"] <- 1
}

# Format data 
occasions$Time <- hour(occasions$start)
occasions$Site <- as.factor(occasions$Site)

# format data for cbind(success, failure)
occasions_cbind <- occasions %>% 
  group_by(Site, Time) %>% 
  summarize(success = sum(capt),
            failure = n() - success)
head(occasions_cbind)

```

We start by fitting a model that describes a *unimodal* activity pattern. This structure includes only the first of the two cosine terms in equation 1 (which translates in the first and second terms of equation 2), and a random intercept and a random slope as described in chapter \@ref(est).   

``` {r univsbim2}
# Unimodal 
unimodal <- mixed_model(fixed = cbind(success, failure) ~ cos(2*pi*Time/24) + sin(2*pi*Time/24), 
                        random = ~ cos(2*pi*Time/24) + sin(2*pi*Time/24) || Site,
                        family = binomial(),
                        data = occasions_cbind)

summary(unimodal)
```

To code for a *bimodal* activity pattern, we extend the model structure above by also including the second cosine term in equation 1 (i.e. third and forth terms in equation 2).

``` {r univsbim3, cache=TRUE, cache.path="cache/"}
# Bimodal 
bimodal <- mixed_model(fixed = cbind(success, failure) ~ cos(2*pi*Time/24) + sin(2*pi*Time/24)+
                                                         cos(2*pi*Time/12) + sin(2*pi*Time/12), 
                       random = ~ cos(2*pi*Time/24) + sin(2*pi*Time/24)+
                                  cos(2*pi*Time/12) + sin(2*pi*Time/12) || Site,
                       family = binomial(),
                       data = occasions_cbind)

summary(bimodal)
```

Finally, we use a model without any cosine term to describe a *cathemeral* activity patterns. We include a random intercept but cannot include a random slope due the structure of model itself. 

``` {r univsbim4}
null_mod <- mixed_model(fixed = cbind(success, failure) ~ 1, 
                        random = ~ 1 | Site,
                        family = binomial(),
                        data = occasions_cbind)
summary(null_mod)
```

**Model selection.** We can now proceed to compare the different model structures and associated hypotheses. We can do this using the Akaike Information Criterion (AIC; @AIC).

``` {r univsbim5}
# AIC comparison
AIC(null_mod, unimodal, bimodal)
```

We can also compare the models using a Likelihood Ratio Test (LRT).

``` {r univsbim6}
lrtest(null_mod, unimodal, bimodal)
```

Both comparisons suggest that the bimodal pattern is the most supported model among those compared. We can, thus, conclude that coyotes have a bimodal activity pattern.

We can also visually compare predicted activity curves based on these models. As in other examples, we first use the function `GLMMadaptive::effectPlotData` to predict the activity curves throughout the 24-hour cycle and then use `ggplot` to plot the results.

``` {r univsbim7}
# build a new dataset 
newdat <- with(occasions_cbind, expand.grid(Time = seq(min(Time), max(Time), length.out = 24)))

# obtain the estimated activity curves
predict_unimodal <- effectPlotData(unimodal, newdat, marginal = FALSE) %>% 
  mutate(Model = "Unimodal")
predict_bimodal <- effectPlotData(bimodal, newdat, marginal = FALSE) %>% 
  mutate(Model = "Bimodal")
predict_cathemeral <- effectPlotData(null_mod, newdat, marginal = FALSE) %>% 
  mutate(Model = "Cathemeral")

# join and plot results
pl_shapes <- rbind(predict_unimodal, predict_bimodal, predict_cathemeral) %>% 
  ggplot(., aes(x = Time, y = plogis(pred), group = Model, fill = Model)) +
  geom_line(aes(colour = Model)) +
  geom_ribbon(aes(ymin = plogis(low), ymax = plogis(upp), colour = NULL), alpha = 0.3) +
  labs(x = "Time of Day (Hour)", y = "Predicted Activity Pattern \n (probability)")+
  theme_minimal()+
  theme(legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size=10,face="bold"),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-5,-10,-10,-10),
        plot.title = element_blank(),
        axis.line = element_line(colour = 'black', linetype = 'solid'),
        axis.ticks = element_line(colour = 'black', linetype = 'solid'),
        axis.title = element_text(size=9,face="bold"),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(colour = 'lightgrey', linetype = 'dashed', linewidth=0.5),
        panel.grid.minor.x = element_blank(),
        strip.text = element_text(size = 9, colour = "black", face = "bold", hjust = 0))+
  #panel.grid.minor.x = element_line(colour='deepskyblue4', linetype = 'dotted', size=0.35))+
  scale_x_continuous(breaks=seq(0,23,length.out=7), labels=seq(0,24,4))
pl_shapes

# save plot
ggsave(plot = pl_shapes, filename = "ms_figures/Fig2_Uni_vs_Bim_comparison.jpg", device = "jpeg", units = "cm", width = 12, height = 10, dpi = 600)

```

We can test other hypotheses related to the shape of the activity curve for our target species (e.g. three or more peaks) by simply adding additional cosine terms to the model structure and then comparing the different structures using AIC or LRT. However, going beyond two peaks might be computationally challenging, especially when a random slope is included. In these cases, one might need to consider reducing model complexity by only including a random intercept.
